---
title: "Processing of BD Vulcan Experiments from FCS/WSP into tables"
author: Daniel Schraivogel
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Processing of BD Vulcan Experiments from FCS/WSP files.}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

Built with `r getRversion()`.

```{r include=FALSE}
#Flow specific packages
library(flowCore) #allows to import FCS files
library(CytoML) # allows to import xml from FlowJo
library(flowWorkspace)

#Other packages
library(tidyverse)
```

Define data to import:

```{r message=FALSE, include=FALSE}
#########################################ENTER PATHNAMES HERE##################################################
wsp <- system.file('extdata', 'FlowJo_file.wsp', package = 'ICS2021')
fcs <- system.file('extdata', 'HeLa_H2B_100.fcs', package = 'ICS2021')
path_to_fiji  <- system.file('extdata/images-processed/00000000', package = 'ICS2021')

## Careful: Doesn't work with elliptical gates (gate is found but no cells can be assigned to gate). Only validated for polygon gates so far.
###############################################################################################################
```
<span style="color:red">**The wsp file and project `r paste0(wsp)` was selected as input!**</span>

Import data via flowCore (to have the correct parameters):

```{r include=FALSE}
fc <- as_tibble(exprs(read.FCS(fcs))) 
fc <- fc %>% mutate(imageid = paste0("id", sprintf("%0.8d", 0:(dim(fc)[1]-1)))) #Add imageid
```

Import via flowWorkspace (to include gating information):

1. get wsp (as done so far)
2. get gates (as done so far)
3. generate complete_gated (as done so far)

```{r include=FALSE}
gs  <- flowjo_to_gatingset(open_flowjo_xml(wsp, sample_names_from="sampleNode"), name = 1) #flowWorkspace package

#Extract event table from flowworkspace
fw <- as_tibble(exprs(gs_pop_get_data(gs)[[1]])) %>% 
  mutate(imageid = paste0("id", sprintf("%0.8d", 0:(dim(fc)[1]-1)))) #Add imageid
```

## Explore flowcore workspace with gating information:

How many cells are associated with different gates? 

```{r}
gs_pop_get_count_fast(gs) #Gating stats from flowWorkspace package
```

Show gating hierarchy of sample.

```{r echo=FALSE}
plot(gs) #Show gating hierarchy as figure
```

These are all the gates present in the gating set:

```{r}
allnodes <- gs_get_pop_paths(gs, path = "auto") # from flowWorkspace package
allnodes2 <- gs_get_pop_paths(gs, path = 1)
allnodes
```

## Generate event table with gating information

Next, a table is created containing all events, and the gating information added to the events as new parameter columns.

```{r include=FALSE}
# Get root which are all events
root <- as_tibble(exprs(gs_pop_get_data(gs, allnodes[1])[[1]])) 
root <- root %>% mutate(imageid = paste0("id", sprintf("%0.8d", 0:(dim(root)[1]-1))))

# Generate table "test" which contained image id and one column with FALSE/TRUE for each gate in allnodes
test <- allnodes[-1] %>% lapply(
  function(x){
      var <- as_tibble(exprs(gs_pop_get_data(gs, x)[[1]])) %>% 
      mutate(new = TRUE)
      
      var <- left_join(root, var) %>% select(imageid, new) %>% mutate(new = ifelse(is.na(new), FALSE, new))
      
      colnames(var)[2] <- x
      
      return(var)
    }
)

# Reduce
reduced <- reduce(test, inner_join)

# Check if two events were absolutely identical (this is very unlikely)
any(duplicated(reduced$imageid))

# Generate complete gated column with the events from flowcore (= correct values) and the gating information from flowworkspace
complete_gated <- fc %>% left_join(reduced, by = "imageid")
```

Add file names and paths to table.

```{r}
#Get tiff file names from images folder, also works if images are in subfolders since recursive is TRUE. 
tiff_dir <- system.file('extdata/images', package = 'ICS2021')
files <- list.files(path = tiff_dir, full.names = F, pattern = ".tiff$", recursive = TRUE)

#CAREFUL, REGULAR EXPRESSION MIGHT NOT BE OPTIMAL IN ALL CASES. Currently, it looks for and 8 ({8}) digit number ([0-9]) starting from the end of the string ($).
image_names <- as_tibble(files) %>% 
  mutate(imageid = paste0("id", str_extract(str_extract(value, "[0-9]{8}.tiff"), "[0-9]{8}"))) %>% 
  mutate(path = paste0("../images/", value)) %>% 
  mutate(abstiffpath = paste0(system.file('extdata', package = 'ICS2021'), "images/", value)) %>% 
  select(-value)

#Leftjoin to big table "complete_gated"
complete_gated <- complete_gated %>% left_join(image_names, by = "imageid") 
```

Plot cell subpopulations for the different populations.

```{r eval=FALSE, include=FALSE}
complete_gated %>% 
  ggplot(aes(x = log10(`SSC-A`), y = log10(`FSC-A`))) + 
  geom_point(alpha = 0.3, size = 0.5) +
  scale_fill_discrete(limits = c('FALSE', 'TRUE'))
```

## Some tiff dataset properties (e.g. image size)

```{r, fig.width = 7}
paths <- list.files(path = system.file('extdata/images/', package = 'ICS2021'), 
                    full.names = T, pattern = ".tiff$", recursive = TRUE)

#Get filesize measurements for tiff files
tiffsize <- as_tibble(file.size(paths))

#Generate table with all measurements
tiffparams <- tibble(Filename = files, filesize = tiffsize$value) %>% 
  mutate(imageid = paste0("id", str_extract(str_extract(Filename, "[0-9]{8}.tiff"), "[0-9]{8}")))
tiffparams$filesize <- round(tiffparams$filesize/1000, 0)
tiffparams <- left_join(tiffparams, complete_gated, by = "imageid")

#Generate stats table
tiffstats <- tiffparams %>% group_by(filesize) %>% summarise(counts = n())
tiffstats <- tiffstats %>% mutate(pct = 100*counts/sum(tiffstats$counts))

#Generate file size figure
filesize_plot <- ggplot(tiffstats, aes(x = filesize, y = pct)) + 
  geom_bar(stat="identity") +
  xlab("File size in KB") +
  ylab("Percentage of events") +
  coord_flip() +
  theme_grey(18) 

filesize_plot

#Add tiff size as parameter to complete_gated
tiffparamsmerge <- tiffparams %>% select("imageid", "filesize")
colnames(tiffparamsmerge) <- c("imageid", "tiffsizekb")
complete_gated <- complete_gated %>% left_join(tiffparamsmerge, by = "imageid")
```

Add jpg paths to table.

```{r}
if(path_to_fiji == ""){
  print("No JPG data available.")
} else {
  #Check if jpeg files available (are generated in Fiji)
  jpg_files <- list.files(path = path_to_fiji, full.names = F, pattern = ".jpg$", recursive = TRUE, include.dirs = TRUE)

  #CAREFUL, REGULAR EXPRESSION MIGHT NOT BE OPTIMAL IN ALL CASES. Currently, it looks for and 8 ({8}) digit number ([0-9]) starting from the end of   the string ($).
  jpg_names <- as_tibble(jpg_files) %>% 
    mutate(imageid = paste0("id", str_extract(str_extract(value, "[0-9]{8}.jpg"), "[0-9]{8}"))) %>% 
    mutate(jpgpath = paste0("../", path_to_fiji, "/", value)) %>% 
    mutate(absjpgpath = paste0(path_to_fiji,  "/", value)) %>% 
    #mutate(jpgpath = paste0(path_to_fiji, value)) %>% 
    select(-value)

  #Leftjoin to big table "complete_gated"
  complete_gated <- complete_gated %>% left_join(jpg_names, by = "imageid")
}
```

Polish tables.
```{r}
#Delete events without images from complete_gated
complete_gated <- complete_gated %>%
  filter(WaveformPresent > 0)

#Generate table for processing in Fiji:
#Differences:
#(1)In case not all images were generated by vulcan (e.g. I only selected a subset of images to process), there will be NAs in the path column (although WaveForm is present). Therefore, generate a second smaller table without those NA columns.
#(2)Skip irrelevant columns to have smaller faster table.
reduced_gated <- complete_gated %>% 
  filter(!is.na(path)) %>% 
  select(c("path", allnodes[-1]))

#complete_gated <- complete_gated %>% mutate(SampleID = unlist(strsplit(dir, "/"))[5])
```
