---
title: "Identifying mitotic phases from image sorting data"
author: "Benedikt Rauscher"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Classification of mitotic phases}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results='hide', warning=F, message=F)
```

# Dependencies

```{r}
library(uwot)
library(caret)
library(ComplexHeatmap)
library(reshape2)
library(tidyverse)
library(cowplot)
```

Set cowplot theme for plotting.

```{r}
theme_set(theme_cowplot())
```

Define function namespace.

```{r}
filter <- dplyr::filter
select <- dplyr::select
rename <- dplyr::rename
```

# Characteriztion of ICS parameter properties

## Figure 2A Triplicate measurements Size Ki67

```{r}
## load data
data('data_2a', package = 'ICS2021')

## make scaled density plot
data_2a %>% ggplot(aes(`Size (NFKB FITC)`, color = rep)) + 
  geom_density(aes(y = ..scaled..)) + 
  scale_y_continuous(expand = c(0, 0))

## .. and boxplot
n_group <- data_2a %>% count(rep) %>% pull(n) %>% min()
data_2a %>% group_by(rep) %>% slice_sample(n = n_group) %>% ungroup() %>% 
  ggplot(aes(rep, `Size (NFKB FITC)`)) + 
  geom_boxplot()
```

## Figure 2B Triplicate measurements RM DyeCycle

```{r}
## load data
data('data_2b', package = 'ICS2021')

## plot scaled density
data_2b %>% select(rep, `Radial Moment (DyeCycle Green FITC)`, `Radial Moment (NFKB FITC)`) %>% 
  gather(var, value, -rep) %>% filter(!is.na(value)) %>% 
  ggplot(aes(value, color = rep)) + 
  geom_density(aes(y = ..scaled..)) +
  scale_y_continuous(expand = c(0, 0))

## and boxplot
n_group <- data_2b %>% count(rep) %>% pull(n) %>% min()
data_2b %>% select(rep, `Radial Moment (DyeCycle Green FITC)`, `Radial Moment (NFKB FITC)`) %>% 
  gather(var, value, -rep) %>% filter(!is.na(value)) %>% 
  group_by(rep) %>% slice_sample(n = n_group) %>% ungroup() %>% 
  ggplot(aes(rep, value)) + 
  geom_boxplot() 
```

## Figure 2C Triplicate measurements of Ecc SSC

```{r}
## load data
data('data_2c', package = 'ICS2021')

## make scaled density plot for eccentricity
data_2c %>% ggplot(aes(`Eccentricity (SSC)`, color = rep)) + 
  geom_density(aes(y = ..scaled..)) + 
  scale_y_continuous(expand = c(0, 0))

n_group <- data_2c %>% count(rep) %>% pull(n) %>% min()
data_2c %>% group_by(rep) %>% slice_sample(n = n_group) %>% ungroup() %>% 
  ggplot(aes(rep, `Eccentricity (SSC)`)) + 
  geom_boxplot()
```

## Figures 2D Triplicate measurements of Golgi MI

```{r}
## load data
data('data_2d', package = 'ICS2021')

## make scatter plot for Golgi MI
data_2d %>% ggplot(aes(`MaxIntensity (NFKB FITC)`, `FITC-A`, color = treatment)) + 
  geom_point() + 
  facet_wrap(~ rep) + 
  panel_border() +
  scale_y_continuous(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0))

## rep 1 only
data_2d %>% filter(rep == 'R1') %>%
  ggplot(aes(`MaxIntensity (NFKB FITC)`, `FITC-A`, color = treatment)) + 
  geom_point() +
  scale_y_continuous(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0))

## box plots
n_group <- data_2d %>% count(rep, treatment) %>% pull(n) %>% min()
data_2d %>% select(c(rep, treatment, `MaxIntensity (NFKB FITC)`, `FITC-A`)) %>% 
  group_by(rep, treatment) %>% slice_sample(n = n_group) %>% ungroup() %>%
  gather(var, value, -rep, -treatment) %>% 
  unite(condition, treatment, rep, remove = F) %>% 
  ggplot(aes(condition, value, fill = treatment)) + 
  geom_boxplot() + 
  facet_wrap(~ var, scales = 'free_y', ncol = 1) + panel_border()

# median jitter plots
data_2d %>% select(c(rep, treatment, `MaxIntensity (NFKB FITC)`, `FITC-A`)) %>% 
  gather(var, value, -rep, -treatment) %>% 
  unite(condition, treatment, rep, remove = F) %>% 
  group_by(condition, rep, treatment, var) %>%
  summarize(median_val = median(value)) %>% ungroup() %>%
  ggplot(aes(treatment, median_val)) + 
  geom_jitter(width = 0.2) + 
  stat_summary(fun = 'mean', fun.min = 'mean', fun.max = 'mean',
               color = 'red', geom = 'crossbar', width = 0.2) + 
  ggsignif::geom_signif(comparisons = list(c('BFA', 'unt')), test = 't.test') +
  facet_wrap(~ var, scales = 'free_y', nrow = 1) +
  panel_border()


## different shapes for different replicates
data_2d %>% ggplot(aes(`MaxIntensity (NFKB FITC)`, `FITC-A`, color = treatment)) + 
  geom_point(aes(shape = rep)) +
  scale_y_continuous(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0))

## 2d density plot?
data_2d %>% ggplot(aes(`MaxIntensity (NFKB FITC)`, `FITC-A`, color = treatment, linetype = rep)) +
  geom_density_2d() +
  scale_y_continuous(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0))
```

## Figure 2J

We repeated the experiment to determine the overlap between the classification by microscopy and the sorted gate 4 times. In Figure 2J we present a heatmap showing the results for one representative experiment. We also provide visualizations that illustrate the reproducbility between the replicates. 

There are two Excel tables each containing data for two replicates. Telophase is a combination of the columns "telophase", "broken/tiny cell condensed" and "G1/tiny cell decondensed".

```{r}
## load data
data('data_2j', package = 'ICS2021')
data('data_pm', package = 'ICS2021')

## make heatmap for each experiment
data_2j %>% group_by(sheet, rep) %>% 
  group_map(~ {
    mat <- .x %>% as.data.frame() %>% column_to_rownames('stage') %>%
      as.matrix() %>% t()
    Heatmap(mat, cluster_rows = F, cluster_columns = F, border = 1,
            col = circlize::colorRamp2(seq(0, 100, by = 20), 
                                       c('#ffffff','#9ecae1','#6baed6','#4292c6','#2171b5','#084594')))
  }) %>% reduce(`+`)

## make a heatmap of average values include 95% ci
data2j_avg <- data_2j %>% bind_rows(data_pm) %>%
  pivot_longer(names_to = 'class', values_to = 'purity', -c(stage, rep, sheet)) %>% 
  mutate(purity = ifelse(is.na(purity), 0, purity)) %>%
  group_by(stage, sheet, rep) %>% mutate(purity = (purity/sum(purity))*100) %>% ungroup() %>%
  group_by(stage, class) %>% summarise(ci95 = mean_se(purity, mult = 2)) %>% ungroup()
data2j_avg <- bind_cols(data2j_avg[,c(1, 2)], data2j_avg$ci95) %>%
  mutate(stage = factor(stage, levels = c('Interphase', 'Prometaphase', 'Prometaphase 2x', 'Metaphase', 'Anaphase', 'Telophase', 'G2M')),
         class = factor(class, levels = c('interphase', 'prophase', 'prometaphase', 'metaphase', 'anaphase', 'telophase', 'apoptosis')))
m1 <- acast(data2j_avg, class ~ stage, value.var = 'y')
m2 <- acast(data2j_avg, class ~ stage, value.var = 'ymin')
m3 <- acast(data2j_avg, class ~ stage, value.var = 'ymax')

## draw heatmap
label_fun = function(j, i, x, y, width, height, fill) {
  grid.text(sprintf("%.1f [%.1f, %.1f]", m1[i, j], m2[i, j], m3[i, j]), 
            x, y, gp = gpar(fontsize = 10))
}
Heatmap(m1, cluster_rows = F, cluster_columns = F, border = 1,
        col = circlize::colorRamp2(seq(0, 100, by = 20), 
          c('#ffffff','#9ecae1','#6baed6','#4292c6','#2171b5','#084594')),
        cell_fun = label_fun
)

## visualize reproducibility
data_2j %>% 
  pivot_longer(names_to = 'class', values_to = 'perc_purity', -c(stage, rep, sheet)) %>% 
  ggplot(aes(stage, perc_purity, color = class)) + 
  geom_jitter(width = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Figure S2

```{r}
## load data
data('data_supp', package = 'ICS2021')

## only use time point 0 and type "RT" for figure s2 
data_fs2 <- data_supp %>% filter(time == '0h', type == 'RT')
```

I make box plots for the features "Correlation (DRAQ5 PerCP / NFKB FITC)", "FITC-A", "FSC-A" and "SSC-A".

```{r}
set.seed(1234)

manuscript_boxplot <- function(data, feature, ylm){
  ## number of cells to plot
  n_cells <- data %>% count(treatment, rep) %>% pull(n) %>% min()
    
  ## prepare data
  df <- data %>% mutate(comp = paste(treatment, rep, sep = '_')) %>%
    arrange(desc(treatment), rep) %>%
    mutate(comp = factor(comp, levels = unique(comp))) %>%
    group_by(comp) %>% slice(1:n_cells) %>% ungroup() 
  
  ## boxplots
  p1 <- df %>% ggplot(aes_string('comp', paste0('`', feature, '`'), fill = 'treatment')) +
    geom_boxplot(position = 'dodge') +
    ylim(ylm) +
    ggtitle(paste('Number of cells:', n_cells))
  
  ## compare medians
  p2 <- df %>% group_by(treatment, rep, comp) %>% 
    summarize(med = median(get(feature))) %>% ungroup() %>% 
    mutate(treatment = factor(treatment, levels = c('unt', 'TNF'))) %>% 
    ggplot(aes(treatment, med)) + 
    geom_jitter(width = 0.1) + 
    ggsignif::geom_signif(comparisons = list(c('unt', 'TNF')), test = 't.test') + 
    ylab(paste(feature, 'median value'))
  
  return(list(box = p1, jitter = p2))
}

## make box plots for each feature
corr_plots <- manuscript_boxplot(data_fs2, "Correlation (DRAQ5 PerCP / NFKB FITC)", c(-1, 1))
fitca_plots <- manuscript_boxplot(data_fs2, "FITC-A", c(0, 1.4e08))
fsca_plots <- manuscript_boxplot(data_fs2, "FSC-A", c(0, 9e07))
ssca_plots <- manuscript_boxplot(data_fs2, "SSC-A", c(0, 3.5e07))
```

## Figure 2E

I make a density plot comparing the correlation values between treatments and replicates.

```{r}
## all replicates
data_fs2 %>%
  ggplot(aes(`Correlation (DRAQ5 PerCP / NFKB FITC)`, 
             color = treatment, linetype = rep)) +
  geom_density(aes(y = ..scaled..)) + 
  scale_y_continuous(expand = c(0, 0))

## only replicate 1
data_fs2 %>% filter(rep == '1') %>%
  ggplot(aes(`Correlation (DRAQ5 PerCP / NFKB FITC)`, 
             color = treatment)) +
  geom_density(aes(y = ..scaled..)) + 
  scale_y_continuous(expand = c(0, 0))
```

## Figure S4B Temperature effect on correlation stability

```{r}
data_supp %>% filter(`Viable single draq5`) %>% 
  group_by(time, treatment, type, rep) %>% 
  summarize(median_cor = median(`Correlation (DRAQ5 PerCP / NFKB FITC)`)) %>% 
  ungroup() %>% unite(group, type, treatment, remove = F) %>% 
  ggplot(aes(time, median_cor, color = type, group = group)) + 
  geom_point() + stat_summary(fun = 'mean', geom = 'line')
```

## Bead sizing experiments

This experiments tests the accuracy of the size paramter as measured by the imaging sorter. Beads of known sizes were processed by the sorter to check if the size parameter correlates with these known sizes.

```{r}
## load data
data('beads_data', package = 'ICS2021')

## correlation diameter and size
beads_data %>% ggplot(aes(`Diameter [µm]`, value)) + 
  geom_point(aes(color = rep)) + 
  geom_smooth(se = F, method = 'lm', formula = y ~ x + I(x^2))
## coefficients for quadratic fit
lm(value ~ `Diameter [µm]` + I(`Diameter [µm]`^2), data = beads_data)

beads_data %>% ggplot(aes(`Diameter [µm]`, value)) + 
  geom_point(aes(color = rep)) + geom_smooth(method = 'lm') +
  ggpubr::stat_cor(method = 'pearson') +
  scale_y_log10()

## correlation beads area and size
beads_data %>% ggplot(aes(`Area [µm^2]`, value)) + 
  geom_point(aes(color = rep)) + geom_smooth(method = 'lm') +
  ggpubr::stat_cor(method = 'pearson')
```

## Figure S4D

We create a box plot showing viability before and after sorting over 3 replicates.

```{r}
## load data
data('data_fs4', package = 'ICS2021')

## add dapi negative cells
dapi_negative <- data_fs4 %>% group_by(treatment, rep, condition) %>%
  summarise(perc_dapi_negative = (sum(live_gfppos)/n()) * 100) %>%
  ungroup()

## make bar plot?
dapi_negative %>% 
  mutate(group = paste(treatment, condition), 
         group = factor(group, levels = c('untreated PreSort', 'TNFa PreSort', 'TNFa Lower_PostSort', 'TNFa Upper_PostSort'))) %>%
  ggplot(aes(group, perc_dapi_negative)) + 
  stat_summary(fun = 'mean', geom = 'bar') +
  geom_jitter(width = 0.1) + 
  ggsignif::geom_signif(comparisons = list(c('untreated PreSort', 'TNFa PreSort'),
                                           c('TNFa PreSort', 'TNFa Lower_PostSort'),
                                           c('TNFa PreSort', 'TNFa Upper_PostSort')),
                        step_increase = 0.25, test = 't.test') +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100)) 
```


# Analysis of annotated images

We manually annotated a total of 100 images for each of the cell cycle phases. The goal of this analysis is to find out which features explain the different phases and how well they can be distinguished by machine learning methods.

We first load the data from disk.

```{r}
data('mitotic_phases', package = 'ICS2021')
```

We can visualize the distribution of each feature to examine how they vary by mitotic phase.

```{r}
## check feature distributions for different phases
mitotic_phases %>% filter(rep == 'rep2') %>%
  ggplot(aes(z, color = phase)) + 
  geom_density() + 
  facet_wrap(~feature, scales = 'free_y')
```

There seem to be clear differences separating the groups for some of the parameters. I create a heatmap based on the z-score distributions to investigate. We can summarize the distributions by creating a heatmap of median values.

```{r}
## use median feature value for aggregation
feature_mat <- mitotic_phases %>%
  mutate(feature = make.names(feature),
         feature = gsub('H2B.mNeonGreen', '', feature),
         feature = gsub('\\.\\.\\.', '..', feature)) %>%
  acast(feature ~ rep + phase, value.var = 'z', fun.aggregate = median)

## heatmap
Heatmap(feature_mat)
```

## UMAP

We can cluster the labeled cells using an unsupervised dimensionality reduction method like UMAP. If the features are informative then the mitotic phases should be separated in the UMAP plot.

```{r}
## data frame containing the normalized data
data_norm <- mitotic_phases %>% filter(rep == 'rep2') %>%
  select(-val) %>% 
  pivot_wider(names_from = 'feature', values_from = 'z')

## remember row annotations and remove metadata columns
labels <- data_norm$phase
image_ids <- data_norm$imageid
data_norm <- data_norm[,!colnames(data_norm) %in% c('imageid', 'phase', 'cell_nr')]

## run umap on subset of the cells (10,000)
set.seed(1234)
umap_res <- umap(data_norm)
  
## add metadata for umap dimensions
umap_res <- as_tibble(umap_res) %>% mutate(phase = labels, image_id = image_ids)

## visualize
umap_res %>% ggplot(aes(V1, V2, color = phase)) + geom_point() +
  xlab('UMAP 1') + ylab('UMAP 2')
```

## Decision tree

Can we create a decision tree to classify the cells based on their cell cycle phase? We define a function that can fit a decision tree on a training data set (70% of all cells) using 10x cross-validation. For tree fitting we use the 'rpart2' method implemented in the 'caret' package.

```{r}
## exclude -h and -w parameters (-a is more commonly used and
## they are highly correlated)
data_ml <- mitotic_phases %>% select(-z) %>% 
  pivot_wider(names_from = 'feature', values_from = 'val')
data_ml <- data_ml[,!grepl('-H|-W', colnames(data_ml))]

## function for fitting trees
fit_tree <- function(df, r, params, classes){
  ## select parameters
  df <- df %>% filter(phase %in% classes, rep == r)
  df <- df[,colnames(df) %in% c('phase', params)]
  df <- df[,!apply(df, 2, function(x) any(is.na(x)))]
  
  ## fix feature names (exclude special characters)
  colnames(df) <- make.names(colnames(df))
  
  ## split into training and test set (70-30)
  idx <- createDataPartition(df$phase, p = 0.7)[[1]]
  training <- df[idx,]
  
  ## learn model, 10x CV
  control <- trainControl(method = 'repeatedcv', number = 10, repeats = 3, savePredictions = T)
  model <- train(
    phase ~ ., 
    data = training,
    method = 'rpart2',
    trControl = control
  )
  return(list(model = model, index = idx, df = df))
}

## fit tree for both parameter sets
phases <- c('anaphase', 'telophase', 'prometaphase', 'interphase', 'metaphase')
models <- list(
  rep1 = list(r = 'rep1', subset = colnames(data_ml)[-(1:4)], classes = phases),
  rep2 = list(r = 'rep2', subset = colnames(data_ml)[-(1:4)], classes = phases),
  rep3 = list(r = 'rep3', subset = colnames(data_ml)[-(1:4)], classes = phases)
  ) %>% map(~ {
    fit_tree(data_ml, .x$r, .x$subset, .x$classes) 
  })

## visualize the resulting trees
rattle::fancyRpartPlot(models$rep1$model$finalModel, cex = .75, 
                       main = 'Rep 1')
rattle::fancyRpartPlot(models$rep2$model$finalModel, cex = .75, 
                       main = 'Rep 2')
rattle::fancyRpartPlot(models$rep3$model$finalModel, cex = .75, 
                       main = 'Rep 3')
```

The results suggest that the different phases can be nicely separated.

To evaluate the model we generate a confusion matrix based on the 30% cells that were not used for training the model.

```{r}
make_confusion_matrix <- function(tree_obj){
  ## get test data (all labeled cells not used for training)
  test_set <- tree_obj$df[-tree_obj$index,]
    
  ## predict classes for test data 
  pred <- predict(tree_obj$model, test_set)
    
  ## generate confusion matrix
  conf_mat <- confusionMatrix(data = pred,
                  reference = factor(test_set$phase,
                                     levels = levels(pred))) 
}

## get performance metrics
conf_mat <- models %>% map(make_confusion_matrix)

## visualize
perf <- conf_mat %>% map_df(~ .x$byClass %>% as_tibble(rownames = 'phase'), .id = 'rep') %>%
  select(rep, phase, Sensitivity, Specificity) %>% 
  pivot_longer(names_to = 'metric', values_to = 'val',
               Sensitivity:Specificity)

## rename phases so they overlap
perf <- perf %>% 
  mutate(phase = gsub('Class: ', '', phase), 
         phase = ifelse(grepl('oval', phase), 'anaphase', 
                 ifelse(phase == 'Interphase_Prophase', 'interphase',
                 ifelse(grepl('arrest', phase), 'prometaphase', phase))), 
         phase = tolower(phase))

## plot replicates separately
perf %>% 
  ggplot(aes(phase, val, fill = metric)) + 
  geom_bar(stat = 'identity', position = 'dodge') +
  facet_wrap(~rep, ncol = 1) + panel_border() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c('#444444', '#aaaaaa')) +
  ylab('value')

## plot replicates together
perf %>% 
  ggplot(aes(phase, val, fill = metric)) + 
  stat_summary(fun = 'mean', geom = 'bar', position = 'dodge') +
  geom_point(aes(color = rep, group = metric), 
             position = position_dodge(width = 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c('#444444', '#aaaaaa')) +
  ylab('value')
```

This shows that we can sort each cell cycle phase with high sensitivity and specificity.

Finally, we can analyze variable importance based on the tree model to see if other parameters (potentially correlated) could have been used for predictions.

```{r}
## parameter subset
feature_imp <- models %>% map_df(~ {
  varImp(.x$model)$importance %>%
    as_tibble(rownames = 'feature') %>%
    mutate(feature = gsub('H2B.mNeonGreen', '', feature),
           feature = gsub('\\.\\.\\.', '..', feature)) %>%
    arrange(desc(Overall)) %>% 
    mutate(feature = factor(feature, levels = rev(feature)))
}, .id = 'rep')

feature_imp %>% 
  ggplot(aes(feature, Overall)) + 
  stat_summary(fun = 'mean', geom = 'bar') +
  geom_point(aes(color = rep)) +
  scale_y_continuous(expand = c(0, 0)) +
  ylab('feature importance') + 
  coord_flip()

feature_imp %>% 
  ggplot(aes(feature, Overall)) + 
  geom_bar(stat = 'identity') +
  scale_y_continuous(expand = c(0, 0)) +
  ylab('feature importance') + 
  coord_flip() + 
  facet_wrap(~ rep, nrow = 1) + panel_border()

## heatmap based on impartance > 0
important_features <- models %>% map(~ {
  varImp(.x$model)$importance %>%
    as_tibble(rownames = 'feature') %>% 
    filter(Overall > 10) %>% pull(feature)
}) %>% unlist() %>% unique()

feature_mat2 <- feature_mat
rownames(feature_mat2) <- make.names(rownames(feature_mat2))
feature_mat2 <- feature_mat2[rownames(feature_mat2) %in% important_features,]

## heatmap
Heatmap(t(scale(t(feature_mat2))))

## separate replicates
Heatmap(feature_mat2, 
        cluster_columns = F,
        column_split = substr(colnames(feature_mat2), 1, 4))

## heatmap with average values
feature_mat_avg <- feature_mat2 %>% 
  as_tibble(rownames = 'feature') %>% 
  pivot_longer(names_to = 'condition', values_to = 'val', -feature) %>% 
  separate(condition, c('rep', 'phase')) %>% 
  group_by(feature, phase) %>% summarize(val = mean(val)) %>% ungroup() %>%
  acast(feature ~ phase, value.var = 'val')
## order cols
feature_mat_avg <- feature_mat_avg[,c('interphase', 'prometaphase', 'metaphase', 'anaphase', 'telophase')]

## heatmap
Heatmap(t(scale(t(feature_mat_avg))), cluster_columns = F, cluster_rows = F)
```

# Session info

```{r}
sessionInfo()
```
